\textbf{Varying coefficients correlated velocity models in complex landscapes with application to narwhal behavioral response to ship exposure and airgun exposure}

### 2024-05-02 Latex/main.tex
new changes to Latex file

%%PACKAGES
\usepackage{hyperref} 
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{bbold}
\usepackage{float}
\usepackage[margin=0.8in]{geometry}
\usepackage[backend=biber,style=authoryear,backref = true]{biblatex}
\usepackage[affil-it]{authblk}
\usepackage{fancyhdr}
\usepackage{comment}
\usepackage{xcolor}
\usepackage{nicematrix}
\usepackage{multirow}
\usepackage{diagbox}
\usepackage{tikz}
\usepackage{yhmath}
\usetikzlibrary{quotes,angles}
\usepackage{pgfplots}
\pgfplotsset{compat = newest}
\pagestyle{fancy}
\fancyhf{}
\fancyhfoffset[L]{1cm} 
\fancyhfoffset[R]{1cm} 
 \setlength{\headheight}{14pt}
\rhead{Greenland narwhal's behavioural responses}
\lhead{\bfseries \today}
\cfoot{\thepage}

@@

\author{Alexandre Delporte%
	\thanks{\texttt{alexandre.delporte@univ-grenoble-alpes.fr}} }
	\affil{Laboratoire Jean Kuntzmann, Université Grenoble-Alpes, France}
\author{Susanne Ditlevsen%
\thanks{\texttt{susanne@math.ku.dk}}}
\affil{Department of Statistics, University of Copenhagen, Denmark}

\author{Adeline Samson%
	\thanks{\texttt{adeline.leclercq-samson@univ-grenoble-alpes.fr}}}
	\affil{Laboratoire Jean Kuntzmann, Université Grenoble-Alpes, France}

@@

To summarize, the main contributions of this paper are
\begin{itemize}
	\item Derivation of a state-space model for the"Rotational Advective Correlated Velocity Model" (RACVM, \ref{eq: RACVM equation}), and addition of this model in the framework of \texttt{smoothSDE} \texttt{R} package, to enable the use of smooth parameters depending on external covariates, and estimation from noisy observations irregularly spaced in time.
	\item Definition of a constrained version of the RCVM for simulation study, where deviations angles from shoreline and distance to shore are used as covariates to constrain the movement within a polygon, and align the velocity with the boundary of the domain.
	\item Application to real narwhal data to assess a behavioural response of the narwhals, using the exposure covariate defined as the inverse of the distance to the sound source.
\end{itemize}

@@

The transition density for the velocity process is such that 
\begin{equation}V(t+\Delta) \vert V(t)=v \sim \mathcal{N}\left( (I_2-\exp(-A\Delta))\mu + \exp(-A\Delta)v, \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) 
\end{equation}

@@

\[
Q(\Delta)=\left( \Delta-2 \frac{\omega \sin(\omega \Delta)-\frac{1}{\tau} \cos(\omega \Delta)}{\frac{1}{\tau^2}+\omega^2 } \exp\left( -\frac{\Delta}{\tau} \right) +\frac{\tau}{2} \left( \frac{\omega^2-\frac{3}{\tau^2}}{\frac{1}{\tau^2}+\omega^2}-\exp\left( -\frac{2\Delta}{\tau}\right)\right) \right) I_2
\]
The location process is not markovian, but the whole process $(x(t), v(t))^\top$ is.
Both components of the location and velocity processes are independent.
and one can check that for $\omega=0$, the formulas that are derived here match the ones of the continuous-time correlated random walk model \cite{johnson_continuoustime_2008}. Figure~\ref{fig: sample RACVM} shows two samples of a RCVM along with the histograms of the empirical velocity norm. Derivations of the formulas given in this paragraph and the next one are detailed in appendix~\ref{section: detailed RACVM computations}.\\

@@

\begin{equation}
	\forall j \in \{1, \cdots, n\}, \quad 
	\left\{ \begin{array}{l}
		V_{j+1}=\mu_j +\exp(-A_j\Delta_j)(V_j-\mu_j)+\zeta_j \\
		X_{j+1}=X_j+\mu_j \Delta_j+A_j^{-1}(I_2-\exp(-A_j\Delta_j))(_Vj-\mu_j)+\xi_j
	\end{array}\right. \\
	\label{eq: RCVM autoregressive formula}
\end{equation}

@@

\[
\left\{
\begin{array}{l}
	Y_j=Z\alpha_j+\varepsilon_j \\
	\alpha_{j+1}=T_j \alpha_j+B_j \mu_j + \eta_j
\end{array}
\right.\]
where the link matrices are 
\[Z=\begin{pmatrix} I_2 & 0_{2,2}\end{pmatrix} \quad  T_j=\begin{pmatrix} I_2 & A_j^{-1}(I_2-\exp(-A_j \Delta_j)) \\ 0_{2,2} & \exp(-A_j \Delta_j) \end{pmatrix} \quad B_j=\begin{pmatrix}
	\Delta_j I_2-A_j^{-1}(I_2-\exp(-A_j\Delta_j)) \\
	I_2-\exp(-A_j\Delta_j)\end{pmatrix}
\]

@@

Unconstrained movement of animals that do not exhibit a clear rotational movement, as it is often the case for marine mammals, is generally modeled through a CVM, that is equation~\ref{eq: RACVM equation} with $\omega=0$.
To get a constrained version of the CVM, it is natural to make use of a RCVM when the process is close to the shore and the velocity is pointing toward the shore to represent how the animals turn in reaction to the shore.
This can be achieved by choosing a smooth function that is zero far away from the shore and non zero close to the shore for $\omega$.
Denote $\mathcal{D} \subset \R^2$ the polygon defining the domain of the process $X$. The new formulation of the equation is 
\begin{equation} \left\{
	\begin{array}{l}
		dX(t)=V(t) dt \\
		dV(t)=-A(X(t),V(t))V(t)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t)

@@

Here, $A(X(t),V(t))=\begin{pmatrix} 
	\frac{1}{\tau} & -\omega(t) \\
	\omega(t) & \frac{1}{\tau}
\end{pmatrix}$
where $\omega(t)=f_{\omega}(D^{shore}(t),\Theta(t))$, $D^{shore}(t)=d(X(t),\partial\mathcal{D})$ is the distance to the boundary of the domain,
$\Theta(t)=\widehat{(X(t)-p(t),V(t))}$, $p(t)$ is the projection of $X(t)$ on $\partial \mathcal{D}$ and $f_{\omega}$ is a smooth function from $\R^{+} \times [-\pi,\pi]$ into $\R$.

Solving explicitly such a model is out of reach due to the non-linearity induced by the distance to the shore and the angle, but it is possible to get an approximate solution.
For any $t \geq 0$ and $\Delta>0$, if $\Delta$ is small enough, we can use the approximations
\[\forall s \in [t,t+\Delta], D^{shore}(s) \simeq D^{shore}(t) \quad \Theta(s)\simeq \Theta(t)\]
In this case, the formulas of section~\ref{section: RACVM} are still valid.
The question still hanging is : what are the conditions on the function $f_{\omega}$ to ensure that the process is constrained ?
We don't formulate a comprehensive answer here, but only give examples of interpretable smooth functions that result in constrained samples of $X$.
An important spatial scale of the rotational movement is $\rho=\frac{\nu}{\omega}$, which is the diameter of rotation. It should always remain significantly lower than $D^{shore}$ for the process to be constrained. Moreover, a rotation needs to be considered only near the boundary of the domain, that is when $D^{shore}$ approaches $0$, and when the narwhal is heading in the direction of the shore, that is for $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$.
The closer we are to the shore, the faster the velocity should rotate,
whereas at a certain distance from the coast, there should be no further effect of the shore, meaning that $\omega$ should be $0$. Far from the coast, the movement model is simply an integrated Ornstein-Uhlenbeck process, as in the unconstrained case \cite{johnson_continuoustime_2008}.
A weight based on the distance to the shore can be defined as follows :
\[c(D^{shore})=\exp\left(-\kappa \times \left(\frac{D^{shore}}{D_0}\right)^2\right)\mbox{ for some } D_0>0 \mbox{ and } \kappa>0\]

@@

This parametrization of $\omega$ is only motivated by the observations above, and do not pretend to be optimal. Any smooth function that meets minimum requirements could be used instead. The general model is very flexible, and different species-specific behaviours could be described. An alternative is to specify the smooth function $f_{\omega}$ using spline basis functions. \\
Figure~\ref{fig: trueomega} shows a smooth function obtained from equation~\ref{eq: smooth omega} for the parameter $\omega$, while figure~\ref{fig: splineomega} represents a smooth function $f_{\omega}$ obtained as a combination of nine cubic spline basis functions. In the sequel, such models with $\omega$ as a smooth function of $D^{shore}$ and $\Theta$ will be refered to as Constrained Rotational Correlated Velocity Models (CRCVM). A sample of a CRCVM with smooth function $f_{\omega}$ defined as y equation~\ref{eq: smooth omega} is illustrated in Figure~\ref{fig: samples_fjords_CRCVM}. The domain in which the process $X$ is constrained represents Scoresby-Sound fjords system coastline in South-East Greenland.

@@

\subsection{Simulation study}

@@

\subsubsection{Simulation of constrained motion}
\label{section: simulation constrained motion}
In order to experiment the constrained models we simulated a CRCVM within different domains for smooth parameters $\omega$ depending on the angle $\Theta$ and the distance to the boundary of the domain $D^{shore}$.\\

100 batches of 6 individual trajectories over $12$ hours with time step $h=1$ min were sampled in a rectangular domain with sides of length $8$ km from a CRCVM. Gaussian noise with standard deviation $\sigma_{obs}=10$ m is considered to represent measurement error and approach a practical case. Denoting $n$ the number of observations for each individual trajectory, $Y_{ij}$ the observation with noise and $X(t_{ij})=X_{ij}$ the true location, the precise statistical set up is the following:

\begin{equation}  \forall  i \in \{1,\cdots,6\}, \quad  \left\{
	\begin{array}{l}
		Y_{ij}=X_{ij}+\varepsilon_{ij}, \quad j \in \{1,\cdots, n\} \\
		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2) \\
		dX_i(t)=V_i(t) dt \quad \\
		dV_i(t)=-A(X_i(t),V_i(t))V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t)  \\
		\log(\tau_i)=\log(\tau_0)+b_{\tau,i}, \\\
		\log(\nu_i)=\log(\nu_0)+b_{\nu,i}, \\
		\begin{pmatrix} b_{\tau,i} \\ b_{\nu,i} \end{pmatrix} \underset{i.i.d}{\sim} \mathcal{N}\left(0,\begin{pmatrix} \sigma_{\tau}^2 & 0 \\ 0 & \sigma_{\nu}^2 \end{pmatrix}\right) \\
		\omega_i(t)=f_{\omega}(D^{shore}_i(t),\Theta_i(t)), \quad t \geq 0\\
		f_{\omega}(D^{shore},\Theta)=\omega_0+\sum_{k=1}^{L} \omega_k \psi_k(D_{shore},\Theta), \quad (D_{shore},\Theta) \in \R^+ \times [-\pi,\pi]
	\end{array}
	\right.
	\label{eq: rect simulation parameters}
\end{equation}
Here, $L=8$ and $\psi_k$ are bivariate basis functions constructed through tensor products of univariate cubic spline basis functions. We refer to \cite{wood_generalized_2017} section $4.
1.8$ construction of tensor splines basis function. We choose $\sigma_{\tau}=0.1$ h and $\sigma_{\nu}=0.3$ km/h to represent the individual variability, wheras $\tau_0$ and $\nu_0$ are fixed respectively to $1$ h and $4$ km/h.
For each trajectory, initial velocity was set to $0$ and initial position was uniformly sampled in the rectangular domain, at least $500$ metres away from the boundary.   Of the $600$ simulated trajectories, none reach the boundary.\\

100 batches of 6 individual trajectories over $24$ hours were also simulated in Scoresby Sound fjords system in which the narwhals move under constraints in our application. The parameters were set as in~\ref{eq: rect simulation parameters}, with only the coefficients $\omega_k$, $k \in \{1,\cdots,L\}$ changed to adapt to this specific domain.
Again, for each trajectory, initial velocity was set to $0$ and initial position was uniformly sampled within the sea, with the requirement that it should be at least $500$ metres away from the shore. Time steps were also kept constant, equal to $1$ min. None of the $600$ trajectories reached land. Note that if time steps are too large, the assumption of constant covariates between observations is very likely to be violated due to the steepness of the shoreline, resulting in samples that reach land very often. For the fjords system, this already happens if we try to sample directly with $5$ min time steps. Constrained samples within the geometry of Scoresby sound fjords are shown in~\ref{fig: samples_fjords_CRCVM}. Moreover, sampling within the fjords takes much more time than within the rectangular domain, since it is necessary to loop over each polygon of the geometry to compute the nearest shore point for each new position.

\subsubsection{Parameters estimation with \texttt{smoothSDE}}

Estimation with $\texttt{smoothSDE}$ relies on the state-space formulation in section~\ref{section: state space model} with the approximation that the parameters are constant on each time step between consecutive observations.
The complete likelihood is computed from the observations $Y_{ij}$ as a by-product of the Kalman filter algorithm \cite{michelot_varying-coefficient_2021}.
Laplace's approximation of the integral of this complete likelihood over the random effects is computed using $\texttt{TMB}$ \cite{kristensen_tmb_2016} and optimization is performed via the \texttt{optim} function in $\texttt{R}$ with the BFGS gradient method. We refer to \cite{wood_generalized_2017} section $6.6$ for the specification of the regularization terms based on a wiggliness measure in the mixed effect models involving gaussian priors for the spline coefficients $\omega_k$, $k \in \{1,\cdots,L\}$. \\

$100$ batches of noisy samples obtained as described in section~\ref{section: simulation constrained motion} are used as data for estimation, both in the rectangular domain and in Scoresby Sound fjords system. We thus obtain $100$ parameters estimate in the two cases. The covariates $D^{shore}$ and $\Theta$ are computed from the observed positions and the smooth parameter $\omega$ is estimated using bivariate tensor splines of these covariates, given by the function \texttt{te} in \texttt{R}. The marginal spline degree of freedom was set to $3$ for each covariate, which corresponds to one single knot in each dimension. Hence, $9$ coefficients (including the intercept) are to be estimated for the smooth function $\omega$, plus the two parameters $\tau$ and $\nu$, and the individual effects standard deviations $\sigma_{\tau}$ and $\sigma_{\nu}$. Measurement error $\sigma_{obs}$ is supposed to be known. 
We fixed initial parameter values to $\tau_0=1$ h, $\nu_0 = 1$ km/h,$\omega_k=0$ for all $k \in \{0,\cdots,L\}$. Figure~\ref{fig: marginal_estimates_CRCVM} shows the results of the estimation of $f_{\omega}$ from one batch of $6$ trajectories for different values of the distance to the shore in the rectangular and the fjords domain.

@@

For each batch of $6$ trajectories, we obtain estimates $\widehat{\log(\tau_0)}^{(k)}$, $\widehat{\log(\nu_0)}^{(k)}$, $\hat{\omega}_l^{(k)}, l \in \{0,\cdots, L\}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$ and compute the root mean squared error and the absolute biais of the estimator. The results a summarized in Tables~\ref{table: estimation performance rect} and~\ref{table: estimation performance fjords}.

\begin{table}[H]
		\centering
	\begin{tabular}{|c|c|c|c|}
		\hline
		Coefficient  & True value & Biais & Root mean squared error  \\
		\hline
		$\log(\tau_0)$  & 0& 0.24 &  \\
		\hline
		$\log(\nu_0)$  & 1.38 & 0.03 &  \\
		\hline
		$\omega_0$  &  -0.19  & 0.80 &  \\
		\hline
		$\omega_1$  & 16.68 & 11.54 &  \\
		\hline
		$\omega_2$  & -3.58 & 2.58 &  \\
		\hline
		$\omega_3$  & 13.07 &  8.74 &  \\
		\hline
		$\omega_4$  & 42.11 & 26.36 &  \\
		\hline
		$\omega_5$  & 13.89 & 8.44 &  \\
		\hline
		$\omega_6$  &  61.53 & 51.45 &  \\
		\hline
		$\omega_7$  & 10.15 & 2.54 &  \\
		\hline
		$\omega_8$  &  12.31 & 4.22 &  \\
		\hline
		$\sigma_{\tau}$ & 0.1 & & \\
		\hline
		$\sigma_{\nu}$ & 0.3 & & \\
		\hline
	\end{tabular}
	\caption{Measures of estimation performance in the rectangular domain}
	\label{table: estimation performance rect}
\end{table}

@@

As measurement error grows, the covariates that are estimated from the observed positions are becoming less accurate, which makes estimation more challenging.

@@

Here, we apply our diffusion models to analyse a behavioral response of the narwhals to ship and seismic ariguns exposure.

@@

\subsection{Estimation without constraints}

First, a simple CVM (see equation \ref{eq: CVM equation}) was used to assess any behavioural response of the narwhals to the ship exposure.

>>>>

%%PACKAGES
\usepackage{authblk}
\usepackage{hyperref} 
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{bbold}
\usepackage{float}
\usepackage[margin=0.8in]{geometry}
\usepackage[backend=biber,style=authoryear,backref = true]{biblatex}
\usepackage{fancyhdr}
\usepackage{comment}
\usepackage{xcolor}
\usepackage{nicematrix}
\usepackage{multirow}
\usepackage{diagbox}
\usepackage{tikz}
\usepackage{yhmath}
\usetikzlibrary{quotes,angles}
\usepackage{pgfplots}
\pgfplotsset{compat = newest}
\pagestyle{fancy}
\fancyhf{}
\fancyhfoffset[L]{1cm} 
\fancyhfoffset[R]{1cm} 
 \setlength{\headheight}{14pt}
\rhead{Greenland narwhal's behavioural responses}
\lhead{\bfseries \today}
\cfoot{\thepage}

@@

\author[1]{Alexandre Delporte}
\author[1]{Adeline Samson}
\author[2]{Susanne Ditlevsen}

\affil[1]{Laboratoire Jean Kuntzmann, Université Grenoble-Alpes, France \authorcr
	\{\tt alexandre.delporte, adeline.samson\}@univ-grenoble-alpes.fr}
\affil[2]{Department of Statistics, University of Copenhagen, Denmark \authorcr
	\tt  susanne@math.ku.dk}

@@

To summarize, the main contributions of this paper are
\begin{itemize}
	\item Derivation of a state-space model for the"Rotational Advective Correlated Velocity Model" (RACVM, \ref{eq: RACVM equation}), and addition of this model in the framework of  the \texttt{R} package \texttt{smoothSDE}, to enable the use of smooth parameters depending on external covariates, and estimation from noisy observations irregularly spaced in time.
	\item Definition of a constrained version of the RCVM for simulation study, where deviations angles from shoreline and distance to shore are used as covariates to constrain the movement within a polygon, and align the velocity with the boundary of the domain.
	\item Application to real narwhal data to assess a behavioural response of the narwhals, using the exposure covariate defined as the inverse of the distance to the sound source.
\end{itemize}

@@

The transition density of the velocity Ornstein-Uhlenbeck process is  
\begin{equation}V(t+\Delta) \vert V(t)=v \sim \mathcal{N}\left( (I_2-\exp(-A\Delta))\mu + \exp(-A\Delta)v, \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) 
\end{equation}

@@

\[
Q(\Delta)=\left( \Delta-2 \frac{\omega \sin(\omega \Delta)-\frac{1}{\tau} \cos(\omega \Delta)}{\frac{1}{\tau^2}+\omega^2 } \exp\left( -\frac{\Delta}{\tau} \right) +\frac{\tau}{2} \left( \frac{\omega^2-\frac{3}{\tau^2}}{\frac{1}{\tau^2}+\omega^2}-\exp\left( -\frac{2\Delta}{\tau}\right)\right) \right) I_2
\]
The location process is not markovian, but the whole process $(X(t), V(t))^\top$ is.
The two components of the location and velocity processes are independent.
and one can check that for $\omega=0$, the formulas that are derived here match the ones of the continuous-time correlated random walk model \cite{johnson_continuoustime_2008}. Figure~\ref{fig: sample RACVM} shows two samples of a RCVM along with the histograms of the empirical velocity norm. Derivations of the formulas given in this paragraph and the next one are detailed in appendix~\ref{section: detailed RACVM computations}.\\

@@

\begin{equation}
	\forall j \in \{1, \cdots, n\}, \quad 
	\left\{ \begin{array}{l}
		V_{j+1}=\mu_j +\exp(-A_j\Delta_j)(V_j-\mu_j)+\zeta_j \\
		X_{j+1}=X_j+\mu_j \Delta_j+A_j^{-1}(I_2-\exp(-A_j\Delta_j))(_Vj-\mu_j)+\xi_j
	\end{array}\right. \\
	\label{eq: RACVM autoregressive formula}
\end{equation}

@@

\begin{equation}
\left\{
\begin{array}{l}
	Y_j=Z\alpha_j+\varepsilon_j \\
	\alpha_{j+1}=T_j \alpha_j+B_j \mu_j + \eta_j
\end{array}
\right.
\label{eq: RACVM state space}
\end{equation}
where the link matrices are 
\[Z=\begin{pmatrix} I_2 & 0_{2,2}\end{pmatrix} \quad  T_j=\begin{pmatrix} I_2 & A_j^{-1}(I_2-\exp(-A_j \Delta_j)) \\ 0_{2,2} & \exp(-A_j \Delta_j) \end{pmatrix} \quad B_j=\begin{pmatrix}
	\Delta_j I_2-A_j^{-1}(I_2-\exp(-A_j\Delta_j)) \\
	I_2-\exp(-A_j\Delta_j)\end{pmatrix}
\]

@@

\subsection{Formulation of the model}

Unconstrained movement of animals that do not exhibit a clear rotational movement, as it is often the case for marine mammals, is generally modeled through a CVM, that is equation~\ref{eq: RACVM equation} with $\omega=0$.
To get a constrained version of the CVM, we propose to make use of a new RCVM that includes how the animals turn in reaction to the shore. To do so, we introduce $\omega$ as a smooth function of the distance to the shore and the angle between the velocity vector and the shore normal vector. 
The smooth function should be non zero close to the shore, that is when a change of direction due to the proximity of the shore is expected, and  zero far away from the shore. 
We denote $\mathcal{D} \subset \R^2$ the polygon defining the domain of the process $X$. The new formulation of the equation is 
\begin{equation} \left\{
	\begin{array}{l}
		dX(t)=V(t) dt \\
		dV(t)=-A(X(t),V(t))V(t)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t)

@@

with $A(X(t),V(t))=\begin{pmatrix} 
	\frac{1}{\tau} & -\omega(t) \\
	\omega(t) & \frac{1}{\tau}
\end{pmatrix}$
where $\omega(t)=f_{\omega}(D^{shore}(t),\Theta(t))$, $D^{shore}(t)=d(X(t),\partial\mathcal{D})$ is the distance to the boundary of the domain,
$\Theta(t)=\widehat{(X(t)-p(t),V(t))}$, $p(t)$ is the projection of $X(t)$ on $\partial \mathcal{D}$ and $f_{\omega}$ is a smooth function from $\R^{+} \times [-\pi,\pi]$ into $\R$.

Solving explicitly such a model is out of reach due to the non-linearity induced by the distance to the shore and the angle $\Theta$, but it is possible to get an approximate solution.
For any $t \geq 0$ and $\Delta>0$, if $\Delta$ is small enough, we can use the approximations
\[\forall s \in [t,t+\Delta], D^{shore}(s) \simeq D^{shore}(t) \quad \mbox{and} \quad  \Theta(s)\simeq \Theta(t)\]
In this case, the formulas of section~\ref{section: RACVM} are still valid.
The question still hanging is: what are the conditions on the function $f_{\omega}$ to ensure that the process is constrained ?
We don't formulate a comprehensive answer here, but only give examples of interpretable smooth functions that result in constrained samples of $X$.
An important spatial scale of the rotational movement is $\rho=\frac{\nu}{\omega}$, which is the diameter of rotation. It should always remain significantly lower than $D^{shore}$ for the process to be constrained. Moreover, a rotation needs to be considered only near the boundary of the domain, that is when $D^{shore}$ approaches $0$, and when the narwhal is heading in the direction of the shore, that is for $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$.
The closer we are to the shore, the faster the velocity should rotate,
whereas at a certain distance from the coast, there should be no further effect of the shore, meaning that $\omega$ should be $0$. Far from the coast, the movement model is simply an integrated Ornstein-Uhlenbeck process, as in the unconstrained case.
A weight based on the distance to the shore can be defined as follows :
\[c(D^{shore})=\exp\left(-\kappa \times \left(\frac{D^{shore}}{D_0}\right)^2\right)\mbox{ for some } D_0>0 \mbox{ and } \kappa>0\]

@@

This parametrization of $\omega$ is only motivated by the observations above, and do not pretend to be optimal. Any smooth function that meets minimum requirements could be used instead. The general model is very flexible, and different species-specific behaviours could be described. An alternative is to specify the smooth function $f_{\omega}$ using spline basis functions. Figure~\ref{fig: trueomega} shows a smooth function obtained from equation~\ref{eq: smooth omega} for the parameter $\omega$, while figure~\ref{fig: splineomega} represents a smooth function $f_{\omega}$ obtained as a combination of nine cubic spline basis functions. In the sequel, such models with $\omega$ as a smooth function of $D^{shore}$ and $\Theta$ will be refered to as Constrained Rotational Correlated Velocity Models (CRCVM). A sample of a CRCVM with smooth function $f_{\omega}$ defined as by equation~\ref{eq: smooth omega} is illustrated in Figure~\ref{fig: samples_fjords_CRCVM}. The domain in which the process $X$ is constrained represents Scoresby-Sound fjords system coastline in South-East Greenland.

@@

@@

\subsubsection{Simulation study}
\label{section: simulation constrained motion}
In order to experiment the constrained models we simulated a CRCVM within different domains for smooth parameters $\omega$ depending on the angle $\Theta$ and the distance to the boundary of the domain $D^{shore}$.\\

100 batches of 6 individual trajectories over $12$ hours with time step $h=1$ min were sampled in a rectangular domain with sides of length $8$ km from a CRCVM. Gaussian noise with standard deviation $\sigma_{obs}=10$ m is considered to represent measurement error and approach a practical case. Denoting $n$ the number of observations, which is the same for each individual trajectory, $Y_{ij}$ the observation with noise and $X(t_{ij})=X_{ij}$ the true location, the precise statistical set up is the following:

\begin{equation}  \forall  i \in \{1,\cdots,6\}, \quad  \left\{
	\begin{array}{l}
		Y_{ij}=X_{ij}+\varepsilon_{ij}, \quad j \in \{1,\cdots, n\} \\
		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2) \\
		dX_i(t)=V_i(t) dt \quad \\
		dV_i(t)=-A(X_i(t),V_i(t))V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t)  \\
		\log(\tau_i)=\log(\tau_0)+b_{\tau,i}, \\\
		\log(\nu_i)=\log(\nu_0)+b_{\nu,i}, \\
		\begin{pmatrix} b_{\tau,i} \\ b_{\nu,i} \end{pmatrix} \underset{i.i.d}{\sim} \mathcal{N}\left(0,\begin{pmatrix} \sigma_{\tau}^2 & 0 \\ 0 & \sigma_{\nu}^2 \end{pmatrix}\right) \\
		\omega_i(t)=f_{\omega}(D^{shore}_i(t),\Theta_i(t)), \quad t \geq 0\\
		f_{\omega}(D^{shore},\Theta)=\omega_0+\sum_{k=1}^{L} \omega_k \psi_k(D_{shore},\Theta), \quad (D_{shore},\Theta) \in \R^+ \times [-\pi,\pi]
	\end{array}
	\right.
	\label{eq: rect simulation parameters}
\end{equation}
Here, $L=8$ and $\psi_k$ are bivariate basis functions constructed through tensor products of univariate cubic spline basis functions. We refer to \cite{wood_generalized_2017} section $4.
1.8$ for the construction of tensor splines basis function. We choose $\sigma_{\tau}=0.1$ h and $\sigma_{\nu}=0.3$ km/h to represent the individual variability, wheras $\tau_0$ and $\nu_0$ are fixed respectively to $1$ h and $4$ km/h. The values of the spline coefficients $\omega_k$, $k \in \{0,\cdots,L\}$ are given in Table~\ref{table: estimation performance rect}. For each trajectory, initial velocity is set to $0$ and initial position is uniformly sampled in the rectangular domain, at least $500$ metres away from the boundary. To simulate the trajectories, we suppose that the value of the parameter $\omega$ is constant on each time step, that is equal to 
\begin{equation}
	\omega_{ij}=f_{\omega}(D^{shore}_{ij},\Theta_{ij}), \quad j  \in \{1,\cdots,n\}
\end{equation}
  where $D^{shore}_{ij}=D^{shore}_i(t_{ij})$ and $\Theta_{ij}=\Theta_i(t_{ij})$. The next position and velocity are then obtained from equation~\ref{eq: RACVM state space}. None of the $600$ simulations reached the boundary. \\

100 batches of 6 individual trajectories over $24$ hours were also simulated in Scoresby Sound fjords system in which the narwhals move under constraints in our application. The parameters were set as in~\ref{eq: rect simulation parameters}, with only the coefficients $\omega_k$, $k \in \{1,\cdots,L\}$ changed to adapt to this specific domain. The exact values of the spline coefficients can be found in Table~\ref{table: estimation performance fjords}.
Again, for each trajectory, initial velocity was set to $0$ and initial position is uniformly sampled within the sea, with the requirement that it should be at least $500$ metres away from the shore. Time steps are also kept constant, equal to $1$ min. None of the $600$ trajectories reached land. Note that if time steps are too large, the assumption of constant covariates between observations is very likely to be violated due to the steepness of the shoreline, resulting in samples that reach land very often. For the fjords system, this already happens if we try to sample directly with $5$ min time steps. Moreover, sampling within the fjords takes much more time than within the rectangular domain, since it is necessary to loop over each polygon of the geometry to compute the nearest shore point for each new position.

Estimation with $\texttt{smoothSDE}$ relies on the state-space formulation in section~\ref{section: state space model} with the approximation that the parameters are constant on each time step between consecutive observations.
The complete likelihood is computed from the observations $Y_{ij}$ as a by-product of the Kalman filter algorithm \cite{michelot_varying-coefficient_2021}.
Laplace's approximation of the integral of this complete likelihood over the random effects is computed using the \texttt{R} package \texttt{TMB}  and optimization is performed via the \texttt{optim} function in $\texttt{R}$ with the BFGS gradient method, with the gradient being calculated by automatic differenciation. We refer to \cite{kristensen_tmb_2016} for more details about the derivation of the multidimensional Laplace's approximation and the gradient computations. Similarly, we refer to \cite{wood_generalized_2017} section $6.6$ for the specification of the mixed effect model for the parameter of the SDE and the regularization terms that are based on a wiggliness measure involving gaussian priors for the spline coefficients $\omega_k$, $k \in \{1,\cdots,L\}$. \\

$100$ batches of noisy samples obtained as described in equation~\ref{eq: rect simulation parameters} are used as data for estimation, both in the rectangular domain and in Scoresby Sound fjords system. We thus obtain $100$ parameters estimate in the two cases. The covariates $D^{shore}$ and $\Theta$ are computed from the observed noisy positions. It is worth noting that as measurement error grows, they are becoming less accurate, which makes estimation more challenging. The smooth parameter $\omega$ is estimated using bivariate tensor splines of these covariates, given by the function \texttt{te} in \texttt{R}. The marginal spline degree of freedom is set to $3$ for each covariate to match the simulated data from equation~\ref{eq: rect simulation parameters}. This corresponds to one single knot in each dimension. Hence, $9$ coefficients, including the intercept, are to be estimated for the smooth function $\omega$, plus the two parameters $\tau$ and $\nu$, and the individual effects standard deviations $\sigma_{\tau}$ and $\sigma_{\nu}$. Measurement error $\sigma_{obs}$ is supposed to be known. 
We fix initial parameter values to $\tau_0=1$ h, $\nu_0 = 1$ km/h, $\omega_k=0$ for all $k \in \{0,\cdots,L\}$. Figure~\ref{fig: marginal_estimates_CRCVM} shows the results of the estimation of $f_{\omega}$ from one batch of $6$ trajectoriesin the rectangular and the fjords domain.
The function $f_{\omega}$  is plotted as a function of $\Theta$ for different fixed values of the distance to the shore, representing movement relatively close and far from the shore.

@@

For each batch of $6$ trajectories, we obtain estimates $\widehat{\log(\tau_0)}^{(k)}$, $\widehat{\log(\nu_0)}^{(k)}$, $\hat{\omega}_l^{(k)}, l \in \{0,\cdots, L\}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$, $k \in \{1,\cdots,100\}$ and compute the root mean squared error and the absolute biais of the estimator. The results are summarized in Tables~\ref{table: estimation performance rect} and~\ref{table: estimation performance fjords}.

\begin{table}[H]
		\centering
	\begin{tabular}{|c|c|c|c|}
		\hline
		Coefficient  & True value & Biais & Root mean squared error  \\
		\hline
		$\log(\tau_0)$  & 0& 0.24 &  0.26\\
		\hline
		$\log(\nu_0)$  & 1.38 & 0.03 & 0.04 \\
		\hline
		$\omega_0$  &  -0.19  & 0.80 & 0.82 \\
		\hline
		$\omega_1$  & 16.68 & 11.54 & 12.32 \\
		\hline
		$\omega_2$  & -3.58 & 2.58 &  3.04\\
		\hline
		$\omega_3$  & 13.07 &  8.74 &  8.84\\
		\hline
		$\omega_4$  & 42.11 & 26.36 &  29.04\\
		\hline
		$\omega_5$  & 13.89 & 8.44 & 8.54 \\
		\hline
		$\omega_6$  &  61.53 & 51.45 &  51.60\\
		\hline
		$\omega_7$  & 10.15 & 2.54 & 3.37 \\
		\hline
		$\omega_8$  &  12.31 & 4.22 & 4.45 \\
		\hline
		$\sigma_{\tau}$ & 0.1 & & \\
		\hline
		$\sigma_{\nu}$ & 0.3 & & \\
		\hline
	\end{tabular}
	\caption{Measures of estimation performance in the rectangular domain}
	\label{table: estimation performance rect}
\end{table}

@@

@@

In this section, we apply our diffusion models to analyse a behavioral response of the narwhals to ship and seismic ariguns exposure.

@@

\subsubsection{Estimation without constraints}

First, a simple CVM was used to assess any behavioural response of the narwhals to the ship exposure.
